---
title: "Variance stabilizing transformation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variance stabilizing transformation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette demonstrates the `vst` function for Variance Stabilizing Transformation (VST) of count data, as well as various useful diagnostic plots.


## Setup

I will use the Bioconductor `airway` dataset as an example.

```{r message=FALSE,warning=FALSE}
library(varistran)
library(Matrix)
library(airway)

# Load example data
data("airway")

# Readable sample and gene names
colnames(airway) <- paste0(airway$dex,rep(1:4,each=2))
rownames(airway) <- rowData(airway)$symbol

counts <- assay(airway, "counts")
```

## Variance Stabilizing Transformation

Filtering is recommended as a first step before using `vst`. Use your favourite filtering method, for example `edgeR::filterByExpr` provides a slightly better approach than what I do here:

```{r}
keep <- rowSums(counts >= 10) >= 4
table(keep)

counts_kept <- counts[keep,,drop=FALSE]
```

We can now apply a VST using `vst`. By default Anscombe's VST for negative binomial data is used. This is like a log2 transformation for large counts, but is a bit different for small counts. The data is also normalized for library size, by default with TMM adjustment. Here I request values that can be interpreted as log2 CPM.

```{r}
lcpm <- vst(counts_kept, cpm=TRUE)
```

A better estimate of the dispersion can be obtained by specifying a design matrix. This ensures real signal in the data is not treated as noise.

```{r}
design <- model.matrix(~ 0 + dex + cell, data=colData(airway))

lcpm <- vst(counts_kept, design=design, cpm=TRUE)
```


## Stability plot

`plot_stability` allows assessment of how well the variance has been stabilized. Ideally this will produce a horizontal line, but counts below 5 will always show a drop off in variance.

```{r}
plot_stability(lcpm, counts_kept, design=design)
```


## Biplot

`plot_biplot` provides a two-dimensional overview of your samples and genes using Principle Components Analysis (similar concept to `plotMDS` in limma).

```{r fig.width=6,fig.height=8}
plot_biplot(lcpm)
```


## Heatmap

`plot_heatmap` draws a heatmap. Specify `n=...` to only show the top genes by span of values.

```{r fig.width=6,fig.height=8}
plot_heatmap(lcpm, n=50)
```


### Shiny report

Varistran's various diagnostic plots are also available as a Shiny app, which can be launched with:

```{r eval=FALSE}
shiny_report(lcpm, counts_kept)
```
